---
title: 'hw02: Analyze PWT with R and tidyverse'
author: |
  | Kenji Sato
  | Kobe University
  | mail@kenjisato.jp
date: '`r format(Sys.time(), "%Y/%m/%d")`'
output:
  bookdown::pdf_document2:
    fig_caption: yes
    includes:
      in_header: asset/latex/custom.sty
    keep_tex: no
    latex_engine: pdflatex
    number_sections: yes
    template: null
    toc: no
  bookdown::html_document2: default
bibliography: 
  - asset/bib/R.bib
  - asset/bib/econ.bib
documentclass: scrartcl
---

```{r setup, include=FALSE}
set.seed(3002)
```

# Overview

## Purpose {-}

To become familiar with R and **tidyverse** and 
to play with the Penn World Table [@pwt].

## Instructions {-}

In this assignment, you will

- clone the assignment repository and make a working branch (eg. `solution` branch);
- solve the problems in Section \@ref(problem); 
- write the solutions in `solution.Rmd` and knit the file; 
- commit `solution.Rmd` and `solution.pdf`; and 
- open a Pull Request.
 
# Set Up

Before you get started, please download the Penn World Table dataset and 
place it in an appropriate directory. You can use the helper script 
I provide. Look at the `R` folder, read throught the code in 
`R/pwt-setup.R` and then execute the following line of code in the console.^[
`source()` function reads the file (R script) passed as the first argument and 
executes the R code written in the file. `"R/pwt-setup.R"` is a string that 
specifies a relative path from your working directory to the file. It assumes
that there is an `R` folder under the working directory, and a file named 
`pwt-setup.R` exists in that `R` folder. If you see an error saying "No such
file or directory," your working directory is different from what I expect or 
you may have mistakenly removed the file. 
]

```{r, eval=FALSE}
source("R/pwt-setup.R")
```

Now you should have PWT dataset on your computer. To load
this dataset in R, I would recommend using `haven::read_dta()` 
function from **haven** package, which comes with **tidyverse**.

```{r}
pwt <- haven::read_dta("~/Data/pwt90.dta")
pwt
```


If you see error saying `Error in loadNamespace(name) : there is no package called ‘haven’` in any of your libraries, please install it by running the following code in the console.^[**haven** is a part of 
the **tidyverse** package family. Notice, however, that `library("tidyverse")` does not load
**haven** automatically. You need to `library("haven")` separately or call functions 
in **haven** with the form of `haven::function_name()` like `haven::read_dta()`.]

```{r, eval=FALSE}
install.packages("tidyverse")
```

In the following, we assume that **tidyverse** is loaded on memory. Do this:

```{r, message=FALSE}
library(tidyverse)
```


You might be worried about the disturbing message that tells you there 
are conflicts of names but you do not have to be.

You see this message because both **dplyr** (loaded with **tidyverse**) 
and **stats** (loaded at start up) packages have functions with identical names. 
You can no longer (in this session) use `filter()` function of the **stats**
package simply with `filter()`, because the name now points to `filter()`
function defined in the **dplyr** package. It does not mean you can never 
use the former function; it does mean that you must use it with its full name 
`stats::filter()`. 

# **dplyr** primer

Table \@ref(tab:pwt90) shows all the variables the table has along with 
short descriptions for the variables.

```{r, message=FALSE, include=FALSE}
pwt_vars <- tibble(name = names(pwt), label = NA)
for (i in seq_along(pwt)) {
  pwt_vars$label[i] <- attr(pwt[[i]], "label")
}
```


```{r pwt90, echo = FALSE}
knitr::kable(pwt_vars, caption = "pwt90.dta")
```

Often times, we do not need all of these variables for analysis. 
To trim away unnecessary data, we will make use of **dplyr**, a package for 
data processing, which comes with **tidyverse**.

Since `pwt90` is too big to learn programming concepts with, let's make a 
smaller toy dataset with `tibble()`.^[`tibble` or `tbl_df` is an extension 
of `data.frame` of base R. Run `vignette("tibble")` for more information.]

```{r}
tbl <- tibble(
  id = letters[1:4],
  salary = 400 + rnorm(4, 0, 50),
  sex = c("M", "M", "F", "F")
)
tbl
```

## `filter`

`filter()` can be used to take rows that satisfy certain conditions. To retrieve 
rows with `salary` more than `400`, you can use the below code.

```{r}
filter(tbl, salary > 400)
```

To retrieve rows that `sex` is `"M"`,

```{r}
filter(tbl, sex == "M")
```

To get rows that `sex` is `"M"` and `salary` is more than `400`,

```{r}
filter(tbl, sex == "M" & salary > 400)
```

To get rows that `sex` is `"F"` or `salary` is less than or equal to`400`,

```{r}
filter(tbl, sex == "F" | salary <= 400)
```


## `select`

To choose clumns, use `select`.

```{r}
select(tbl, id, salary)
```

You can remove columns by appending negative sign.

```{r}
select(tbl, - salary)
```

## `mutate` and `transmute`

To manipulate data in columns, use `mutate` or `transmute`.

`mutate` adds new columns. Let's suppose that `salary` is measured in million yen unit 
and that we want to change the unit to thousand yen. This is achieved with the 
following code.

```{r}
mutate(tbl, salary_in_thousand = 1000 * salary)
```

`transmute` removes all variable other than those explicitly specified.

```{r}
transmute(tbl, id, salary_in_thousand = 1000 * salary)
```

## `%>%`

You can combine the above functions (and many others) with pipe operator `%>%` 
from **magrittr** package, on which **dplyr** depends. 

Let's see an example.

```{r}
tbl %>% 
  filter(salary > 400) %>% 
  select(id, sex)
```

This is equivalent to the following.

```{r}
tbl_tmp <- filter(tbl, salary > 400)
select(tbl_tmp, id, sex)
```

Piping makes a chain of commands look much neater.


## `group_by` and `aggregate`

Another operation we might want to perform is to compute group-wise
statistics. The following code computes the ration of the highest salary
to the lowest within each of male and female groups. 

```{r}
tbl %>% 
  group_by(sex) %>% 
  summarise(mean = max(salary) / min(salary))
```


# PWT and plotting with **ggplot2**

Now is the time to work with PWT. Let's focus on the following ten countries.

```{r}
countries <- c("United States", "United Kingdom", "Germany", "France", 
               "Italy", "Japan", "Canada", "China", "Republic of Korea", "India")
```

We extract `country`, `year`, `rgdpo`, `pop`.


```{r}
pwt10 <- 
  pwt %>% 
  filter(country %in% countries) %>% 
  select(country, year, rgdpo, pop)
pwt10
```


To visualize the GDP growth of these countries, we use **ggplot2** package, 
which again comes with **tidyverse**. The following code produces Figure \@ref(fig:rgdp).

```{r rgdp, fig.cap="Real GDP"}
ggplot(pwt10) + geom_line(aes(x = year, y = rgdpo, color = country))
```

The following code produces Figure \@ref(fig:logrgdp). 
The graphs show roughly constant growth of log real GDP. 

```{r logrgdp, fig.cap="Real GDP on log scale"}
ggplot(pwt10) + geom_line(aes(x = year, y = rgdpo, color = country)) + 
  scale_y_log10()
```

# Problem

Consider the period between 1960 and 2014. Compute the 
average annual real GDP growth rates for these countries 
chosen earlier. Which country did grow the fastest?

How about the growth rates for real GDP per capita?

Write your answer along with code in `solution.Rmd`, knit it, and submit 
through a PR.

# References {-}

